The Evolution of Phenotypically Invariant Genetic Networks

### 2017-07-31 handnotes.tex
metabolic example with Kalman decomp

  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). The organism may have a gene regulatory network $\mathcal{S}$ that can depoly a situation specific metabolic strategy. That is depending on both $u{2}$ and $u_{1}$ the organism synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.

@@

  We can see that changing coordinates on this system will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype.

  DO KALMAN EXAMPLE HERE.

%    \subsection*{Metabolic Network}
%  \begin{align*}
%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
%    \end{align*}
%    \begin{align*}
%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
%    \end{align*}
%
%    \begin{align*}
%      CV &= C \\
%      VB &= B \\
%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
%    \end{align*}
%    The GAL regulatory network in \emph{S. cerevisiae} is modelled here. The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose. The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes. This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
%
%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1. $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations. $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
%
%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant. These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable. Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
%
%    \subsection*{Gap Gene Network}

>>>>

  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). The organism may have a gene regulatory network $\mathcal{S}$ that can depoly a situation specific metabolic strategy. That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.

@@

  We can see that changing coordinates on this system (with any $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.

  $\widehat{\mathcal{S}}$ is $4$-dimensional biological system,  \begin{align*}
    \widehat{\mathcal{S}} = \left\{ \begin{array}{ll}
      \widehat{A} = V \begin{bmatrix} X_{1} & X_{2} \\ 0 & A_{\min} \end{bmatrix} V^{-1}; \quad \widehat{B} = V \begin{bmatrix} X_{3} \\ B_{\min} \end{bmatrix}; \quad \widehat{C} = \begin{bmatrix} 0 & C_{\min} \end{bmatrix} V^{-1} \end{array} \right\}
  \end{align*}
     where $V$ is any invertible $4 \times 4$ matrix and $X_{j}$ is any $2 \times 2$ matrix.

     So for example, some system can be wired as follows, and still be input-output equivalent to the minimal metabolic system $A_{\min}$:
     \begin{align*}
       A' &= \begin{bmatrix}
         1.6923  &  1.5385 &  -2.6154 &  -2 \\
         0.8462  &  0.7692 &  -1.3077 &  -1 \\
         1.2692  &  1.1538 &  -1.9615 &  -1.5 \\
         0.4231  &  0.3846 &  -0.6538 &  -0.5
       \end{bmatrix} \\
       B' &= \begin{bmatrix} 4 & 4 \\ 2 & 2 \\ 3 & 3 \\ 1 & 3 \end{bmatrix} \\
         C' &= \begin{bmatrix}  0.2692  &  0.1538  &  0.0385 &  -0.5 \\
         -0.4231 &  -0.3846 &   0.6538  &  0.5 \end{bmatrix}
     \end{align*}

     Despite the present example consisting of a minimal $2 \times 2$ and a non-minimal $4 \times 4$ system, any $n$-dimensional system can be constructed using this method -- applying a change of coordinates to the Kalman decomposition -- to construct a mechanistically different system with identical phenotypic dynamics. Depending on the specifics of a system being modeled, one may have to take care to restrict the free parameter values and network architecutres to be biologically appropriate. 
%    \subsection*{Metabolic Network}
%  \begin{align*}
%    \dot{x}(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \\ 0 & 0 & -1 & -1 \\ -1 & 1 & 0 & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix} \begin{bmatrix} \text{GAL3} \\ \text{GAL4} \\ \text{GAL80} \\ \text{MIG1} \end{bmatrix}(t) \\ &+ \begin{bmatrix} 1 & 0 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 0 \\ 0 & 1 & 1 \end{bmatrix} \begin{bmatrix} \text{galactose} \\ \text{glucose} \\ \text{BTM} \end{bmatrix}(t) \\
%        y(t) &= \begin{bmatrix} 0 & 1 & 0 & -1 \end{bmatrix} \vec{x}(t)
%    \end{align*}
%    \begin{align*}
%      H(z) &= \begin{bmatrix} 1 \\ z^{2} + z + 1 \\ z + 1 \end{bmatrix} \frac{1}{z^{3} + z -1}
%    \end{align*}
%
%    \begin{align*}
%      CV &= C \\
%      VB &= B \\
%      V :&= \begin{bmatrix} 1 & 0 & a & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & b & 0 \\ 0 & 0 & 0 & 1 \end{bmatrix} , V^{-1} = \begin{bmatrix} 1 & 0 & \frac{-a}{b} & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & \frac{1}{b} & 0 \\ 0 & 0 & 0 & 1\end{bmatrix} \\
%        A(a,b) :&= VAV^{-1} = \begin{bmatrix} -a & a+1 & \frac{a^{2}}{b} & -1 \\ 0 & 0 & \frac{-1}{b} & -1 \\ -b & b & a & 0 \\ 0 & 0 & 0 & 0 \end{bmatrix}
%    \end{align*}
%    The GAL regulatory network in \emph{S. cerevisiae} is modelled here. The activating transcription factor GAL4, which is regulated by the presence of galactose and two other TFs, binds to the promoter region of the GAL regulon -- a cluster of genes regulated by the same promoter encoding 3 enzymes (GAL1, GAL7, and GAL10), for the metabolism of the sugar galactose. The repressing transcription factor MIG1, which is activated by the presence of glucose and other proteins omitted here, also binds to the regulon, preventing the expression of the galactose metabolizing enzymes. This network is one of the most studied gene regulatory networks in yeast, and experiments have already demonstrated significant transcriptional variation among different species and genuses of yeast.
%
%    In the present model, the $A$ matrix encodes the interactions among GAL3, GAL4, GAL80, and MIG1. $B$ interprets the input: the quantity of glucose, galactose, and the basal transcription (BTM) rates impact of TF concentrations. $C$ represents the promoter region of the GAL regulon and is the sum of the influence of GAL4 (activating) and MIG1 (repressing).
%
%    $A(a,b)$ represent alternative network structures that produce identical outputs, with identical input and output transformations -- meaning the regulatory contributions of the sugars and the basasl machinery, as well as the promoter of the GAL regulon are held constant. These interactions may be realized in nature, if the regulatory differences predicted by $A(a,b)$ are biologically/mutationally realizable. Mutations influencing the interactions between GAL4 and GAL80 have been experimentally demonstrated \citep{li2010alterations}.
%
%    \subsection*{Gap Gene Network}
